Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-11 pkg/disttree/inst/draft/draft.Rnw
more work on intro, cleanup, and some comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2009 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e8c8925f6162dab52e04413de746c6671c303ade

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("ggplot2")
theme_set(theme_bw(base_size = 18))
library("disttree")
library("Formula")  # FIX ME: should be imported in disttree
library("gamlss")
library("randomForest")
library("lattice")
library("crch")
library("latex2exp")
library("parallel")
library("gamlss.cens")
gen.cens(NO, type = "left")
library("mgcv")
library("MASS")
library("countreg")

@@

\Abstract{
While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to model all distribution parameters. In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting. In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{stasinopoulos2005}) 
each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects. A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{loh2011classification}) 
or random forests. %(\citealt{breiman2001random}). 
%%%<TH> "random forest" generisch; Breiman and Cutler's randomForest </TH>
This class of tree and forest models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods. In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests. 
%%%<TH> break-up very long sentence </TH>
Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.

%For example in the field of probabilistic weather forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) provide common and widely used tools allowing for clear interpretations. Usually acting on the assumption of a normally distributed response variable both parameters, the mean and the variance, depend linearly on covariates in this type of model. However, the restriction to model only linear effects requires further extensions which allow for more flexibility. Generalized additive models for location, scale, and shape (GAMLSS, \citealt{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modeled separately by its own generalized additive model (GAM). A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees (\citealt{loh2011classification}) or random forests (\citealt{breiman2001random}). This class of models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods. We are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests. Applying these novel approaches to precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.

%In regression analysis one is interested in the relationship between a dependent variable and one or more explanatory variables. Various methods to fit statistical models to the data set have been developed, starting from linear models considering only the mean of the response variable and ranging to probabilistic models where all parameters of a distribution are fit to the given data set.\\
%If there is a strong variation within the data it might be advantageous to split the data first into more homogeneous subgroups based on given covariates and then fit a local model in each subgroup rather than fitting one global model to the whole data set. This can be done by applying regression trees and forests.\\
%Both of these two concepts, parametric modeling and algorithmic trees, have been investigated and developed further, however, mostly separated from each other. Therefore, our goal is to embed the progress made in the field of probabilistic modeling in the idea of algorithmic tree and forest models. In particular, more flexible models such as GAMLSS (\citealt{stasinopoulos2005}) should be fitted in the nodes of a tree in order to capture location, scale, and shape as well as censoring, tail behavior etc. while non-additive effects of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree. In distributional forests an ensemble of distributional trees is built and used to calculate weights which are then included in the fitting process of a distributional model. These forest models can detect smooth effects as well as abrupt changes and interactions in a reasonable way without needing any kind of variable selection or information about the expected effects advance and therefore offers a good compromise particularly in complex settings. 
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}

@@

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogenous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects where it would be challenging
to specify a GAMLSS.

In weather forecasting, these properties are appealing in mountainous regions and complex terrain
where small-scale effects are not well-resolved in the NWP outputs. As these are computed at
a coarser scale, there may be abrupt local effects in the meteorological observations within a single
NWP grid cell. To illustrate this in practice, precipitation forecasts are obtained with distributional
forests at 85 meteorological stations in a mountainous region in the Alps, covering mainly Tyrol,
Austria, and adjacent areas. More specifically, a zero-censored Gaussian distribution is
employed to model 24-hour total precipitation so that the zero-censored point mass describes the
probability of no precipitation on a given day (see Figure~\ref{axams_onestation}). Forecasts
for July are established based on data from the same month in the years 1985--2012 along with
a large number of NWP ensemble outputs as covariates. As Figure~\ref{axams_onestation} shows,
the station-wise models yield a full distributional forecast for each day -- here for the same
day (July~24) at one station (Axams) over four years (2009--2012) -- based on the previous 24~years as training data.
The corresponding observations conform reasonably well with the predictions.

@@

In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std. deviation, ensemble minimum and ensemble maximum. All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std. deviation of mean/minimal/maximal values respectively. Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82 covariates are available in this setting.

@@

@@

%%%<TH> always us the term "distributional forest"
As shown in Table~\ref{model_specification} the distributional forest includes all 82 covariates. A single tree model allows to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable. In a distributional forest though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree. To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied. Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation is the most influential variable. In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.

@@

\section*{Computational details}
The corresponding implementation to the novel methods presented in this article is available on R-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}). The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree. Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a distributional forest. All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).

>>>>

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("disttree")
library("Formula")  # FIX ME: should be imported in disttree
library("gamlss")
library("crch")
library("gamlss.cens")
gen.cens(NO, type = "left")
library("mgcv")
library("countreg")

@@

\Abstract{
While many regression models consider only the mean of the response variable recent advances in parametric modeling have made it possible to model all distribution parameters. In that way the underlying distribution can be fully specified allowing for a wide range of analysis and prediction methods which are particularly valuable in applications such as probabilistic weather forecasting. In generalized additive models for location, scale, and shape %(GAMLSS, \citealt{stasinopoulos2005}) 
each distribution parameter can be modeled separately by its own generalized additive model (GAM) capturing linear and non-linear effects. A different approach that would make it possible to also model non-additive effects and interactions is to apply tree-structured models such as regression trees %(\citealt{loh2011classification}) 
or random forests. %(\citealt{breiman2001random}). 
%%%<TH> "random forest" generisch; Breiman and Cutler's randomForest </TH>
This class of tree and forest models will now be integrated in probabilistic forecasting providing a novel general distributional framework for these methods. In particular, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests. 
%%%<TH> break-up very long sentence </TH>
Applying these novel approaches to precipitation data in a mountainous region with a very large number of numerical weather prediction quantities shows one of their main advantages in selecting variables automatically and illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge and experience from meteorology or based on a computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}

@@

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogenous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects where it would be challenging
to specify a distributional regression model like GAMLSS.

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas. More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
82~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{gneiting2005calibrated}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{gamboostLSS2016} using non-homogenous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
82~available regressors.

@@

In particular, 14 basic forecast variables are considered with up to 12 variations of each of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions, the possible variations include the mean, the standard deviation, the minimum and the maximum over all ensemble members, noted as ensemble mean, ensemble std. deviation, ensemble minimum and ensemble maximum. All of these values can be calculated on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour time window.\\
Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standard deviation, noted as ensemble mean of mean/minimal/maximal values or ensemble std. deviation of mean/minimal/maximal values respectively. Table~\ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column. Altogether 82~covariates are available in this setting.

@@

\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAM. Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}

\fixme{A description of the boosting approach is currently missing completely.}

@@

%%%<TH> always us the term "distributional forest"
As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates. A single tree model allows to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable. In a distributional forest though, for each of the trees a subset of these covariates is chosen randomly as possible split variables which is one of the reasons leading to different splitting variables being chosen in each tree. To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied. Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 year are listed in Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation is the most influential variable. In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.

@@

\section*{Computational details}
The corresponding implementation to the novel methods presented in this article is available on \textsf{R}-Forge in an \textsf{R} package called \textbf{disttree} within the project \textbf{partykit} (\url{https://R-Forge.R-project.org/projects/partykit/}). The function \code{disttree} uses either the \code{mob} or \code{ctree} function as a framework to build a tree. Similarly, the function \code{distforest} is based on the \code{cforest} function which provides a framework to build a distributional forest. All three functions, \code{mob}, \code{ctree} and \code{cforest}, are provided in the package \textbf{partykit} (\citealt{hothorn2015partykit}).

\fixme{Avoid nocite, make all relevant citations properly in the text.}
